---
title: "serosv: model infectious disease parameters from serosurveys"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  out.width = "100%",
  dpi = 300,
  warning = F,
  fig.width = 5,
  fig.height = 5
)
library(serosv)
```

Here we demonstrate how to use `serosv` to estimate the seroprevalence and force of infection using some built-in datasets.

## Polynomial models

Use `polynomial_model()` to fit a polynomial model.

We will use the Hepatitis A data from Belgium 1993–1994 for this example.

```{r}
a <- hav_bg_1964
neg <- a$tot -a$pos
pos <- a$pos
age <- a$age
tot <- a$tot
```

Muench's model can be estimated by either defining `k = 1` (a degree one linear predictor) or setting the `type = "Muench"`.

```{r}
muench1 <- polynomial_model(age, pos, neg, k = 1)
summary(muench1$info)

muench2 <- polynomial_model(age, pos, neg, type = "Muench")
summary(muench2$info)
```

We can plot any model with the `plot()` function.

```{r}
plot(muench2)
```

Similarly, we can estimate Griffith's model.

```{r}
gf_model <- polynomial_model(age, pos, neg, type = "Griffith")
plot(gf_model)
```

And Grenfell and Anderson's model.

```{r}
grf_model <- polynomial_model(age,pos,neg,type = "Grenfell")
plot(grf_model)
```

## Farrington's model

Use `farrington_model()` to fit a Farrington's model.

```{r}
rubella <- rubella_uk_1986_1987
rubella$neg <- rubella$tot - rubella$pos

farrington_md <- farrington_model(
   rubella$age, rubella$pos, rubella$tot,
   start=list(alpha=0.07,beta=0.1,gamma=0.03)
   )
plot(farrington_md)
```

## Weibull model

Use `weibull_model()` to fit a Weibull model.

```{r}
hcv <- hcv_be_2006[order(hcv_be_2006$dur), ]
dur <- hcv$dur
infected <- hcv$seropositive

wb_md <- weibull_model(
   t = dur,
   spos = infected
   )
plot(wb_md)
```

## Local estimation by polynomials

```{r}
mump <- mumps_uk_1986_1987
a = age <- mump$age
pos <- mump$pos
tot <- mump$tot
y <- pos/tot
```

Use `plot_gcv()` to show GCV curves for the nearest neighbor method (left) and constant bandwidth (right).

```{r, fig.width=5, fig.height=2.5}
plot_gcv(
   age, pos, tot,
   nn_seq = seq(0.2, 0.8, by=0.1),
   h_seq = seq(5, 25, by=1)
 )

```

Use `lp_model()` to fit a local estimation by polynomials.

```{r}
lp1 <- lp_model(age, pos, tot, kern="tcub", nn=0.7, deg=2)

plot(lp1)
```
