---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%",
  dpi = 300
)
```

# serosv <img src="logo.svg" align="right" width="120"/>

<!-- badges: start -->

[![Codecov test coverage](https://codecov.io/gh/OUCRU-Modelling/serosv/branch/main/graph/badge.svg)](https://app.codecov.io/gh/OUCRU-Modelling/serosv?branch=main) [![R-CMD-check](https://github.com/OUCRU-Modelling/serosv/actions/workflows/R-CMD-check.yaml/badge.svg)](https://github.com/OUCRU-Modelling/serosv/actions/workflows/R-CMD-check.yaml)

<!-- badges: end -->

`serosv` is an easy-to-use and efficient tool to estimate infectious diseases parameters (seroprevalence and force of infection) using serological data. The current version is mostly based on the book "Modeling Infectious Disease Parameters Based on Serological and Social Contact Data -- A Modern Statistical Perspective" by [Hens et al., 2012 Springer](https://link.springer.com/book/10.1007/978-1-4614-4072-7).

## Installation

You can install the development version of serosv with:

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("OUCRU-Modelling/serosv")
```

## Feature overview

### Datasets

`serosv` contains 15 built-in serological datasets as provided by [Hens et al., 2012 Springer](https://link.springer.com/book/10.1007/978-1-4614-4072-7). Simply call the name to load a dataset, for example:

``` r
rubella <- rubella_uk_1986_1987
```

### Methods

The following methods are available to estimate seroprevalence and force of infection.

Parametric approaches:

-   Frequentist methods:

    -   Polynomial models:

        -   Muench's model
        -   Griffiths' model
        -   Grenfell and Anderson's model

    -   Nonlinear models:

        -   Farrington's model
        -   Weibull model

    -   Fractional polynomial models

-   Bayesian methods:

    -   Hierarchical Farrington model

    -   Hierarchical log-logistic model

Nonparametric approaches:

-   Local estimation by polynomials

Semiparametric approaches:

-   Penalized splines:

    -   Penalized likelihood framework

    -   Generalized Linear Mixed Model framework

## Demo

### Fitting rubella data from the UK

Load the rubella in UK dataset.

```{r}
library(serosv)
```

Find the power for the best second degree fractional polynomial with monotonicity constraint and a logit link function. The power appears to be (-0.9,-0.9).

```{r}
rubella <- rubella_uk_1986_1987

best_2d_mn <- find_best_fp_powers(
  rubella,
  p=seq(-2,3,0.1), mc = T, degree=2, link="logit"
)

best_2d_mn
```

Finally, fit the second degree fractional polynomial.

```{r}
fpmd <- fp_model(
  rubella,
  p=c(-0.9, -0.9), link="logit")

plot(fpmd)
```

### Fitting Parvo B19 data from Finland

```{r}
library(dplyr)
parvob19 <- parvob19_fi_1997_1998

# for linelisting data, either transform it to aggregated
transform_data(
  parvob19$age, 
  parvob19$seropositive,
  stratum_col = "age") |>
  polynomial_model(type = "Muench") |>
  plot()

# or fit data as is
parvob19 |>
  rename(status = seropositive) |>
  polynomial_model(type = "Muench") |>
  plot()
```
