---
title: "Model selection"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model selection}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, output=FALSE, warning=FALSE, message=FALSE}
library(serosv)
library(dplyr)
library(magrittr)
```

```{r, warning=FALSE}
data <- parvob19_fi_1997_1998[order(parvob19_fi_1997_1998$age), ] 

aggregated <- transform_data(data$age, data$seropositive, stratum_col = "age")
```

## Generate models comparison `data.frame`

Function `compare_models()` is used for quickly computing comparison metrics for a set of models on a given dataset.

The function takes the following arguments:

-   `data` the dataset that will be fitted to the models

-   `method` the method to generate comparison metrics. It can be the name of one of the built-in methods, or a user-defined function.

-   `…` functions that take a data and return a fitted `serosv` model.

It will then return a `data.frame` with the following columns

-   `label` model identifier. Either user defined name or index based on the order provided.

-   `type` type of model (a `serosv` model class)

-   metrics depending on the method selected

**Built-in `method`**

`serosv` currently provide 2 built-in metrics generating methods

-   `"AIC/BIC"` which returns fitted model's AIC, BIC and Log-likelihood (where applicable)

-   `"CV"` which split the data into "train" and "test" set then return logloss and AUC from model's prediction on the test set

**Sample usage**

```{r message=FALSE}
# ----- Return AIC, BIC ----- 
aic_bic_out <- compare_models(
  data = data %>% rename(status=seropositive),
  method = "AIC/BIC",
  griffith = ~polynomial_model(.x, k=3),
  penalized_spline = penalized_spline_model,
  farrington = ~farrington_model(.x, start=list(alpha=0.07,beta=0.1,gamma=0.03)),
  local_polynomial = lp_model # expect to not return any values
  ) %>% suppressWarnings()

# ----- Return Cross-validation metrics ----- 
cv_out <- compare_models(
    data %>% rename(status=seropositive),
    method = "CV",
    griffith = ~polynomial_model(.x, k=3),
    penalized_spline = penalized_spline_model,
    farrington = ~farrington_model(.x, start=list(alpha=0.07,beta=0.1,gamma=0.03)),
    local_polynomial = lp_model 
  ) %>% suppressWarnings()

aic_bic_out
cv_out
```

With aggregated data

```{r message=FALSE}
# ----- Return AIC, BIC ----- 
aic_bic_out <- compare_models(
  data = aggregated,
  method = "AIC/BIC",
  griffith = ~polynomial_model(.x, k=3),
  penalized_spline = penalized_spline_model,
  farrington = ~farrington_model(.x, start=list(alpha=0.07,beta=0.1,gamma=0.03)),
  local_polynomial = lp_model # expect to not return any values
  ) %>% suppressWarnings()

# ----- Return Cross-validation metrics ----- 
cv_out <- compare_models(
    data = aggregated,
    method = "CV",
    griffith = ~polynomial_model(.x, k=3),
    penalized_spline = penalized_spline_model,
    farrington = ~farrington_model(.x, start=list(alpha=0.07,beta=0.1,gamma=0.03)),
    local_polynomial = lp_model 
  ) %>% suppressWarnings()


aic_bic_out
cv_out
```
