---
title: "Model multivariate serological data"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model multivariate serological data}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, output=FALSE}
library(serosv)
```

## Bivariate Dale model

Fitting bivariate Dale model using `bivariate_dale_model()`

We will use `rubella_mumps_uk` data for this example

```{r}
data <- rubella_mumps_uk
model <- bivariate_dale_model(age = data$age, y = data[, c("NN", "NP", "PN", "PP")], monotonized=TRUE)
```

plot function for `bivariate_dale_model` object takes 3 additional parameters:

-   `y1` and `y2` are labels for 2 diseases being modeled

-   `plot_type` is for the type of plot being returned

    -   `"ci"` returns fitted line for marginal prevalence and its 95% confidence interval

    -   `"sp"` returns conditional and joint seroprevalence and FOI

    -   other inputs would returns all individual plots

```{r, fig.width=7, fig.height=5}
plot(model, y1 = "Rubella", y2 = "Mumps", plot_type = "sp")
```

For line listing data `vzv_parvo_be` , use `generate_quad_matrix()` function to create the (NN, NP, PN, PP) for fitting the model

```{r, eval = F}
generate_quad_matrix(vzv_parvo_be, vzv_res, parvo_res, age, discrete_age = F)
```
