---
title: "Parametric models"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric models}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

```{r setup, output=FALSE}
library(serosv)
```

## Polynomial models

Refer to `Chapter 6.1.1`

Use `polynomial_model()` to fit a polynomial model.

We will use the `Hepatitis A` data from Belgium 1993--1994 for this example.

```{r}
a <- hav_bg_1964
neg <- a$tot -a$pos
pos <- a$pos
age <- a$age
tot <- a$tot
```

### Muench model

**Muench**'s model can be estimated by either defining `k = 1` (a degree one linear predictor) or setting the `type = "Muench"`.

```{r}
muench1 <- polynomial_model(age, pos, neg, k = 1)
summary(muench1$info)

muench2 <- polynomial_model(age, pos, neg, type = "Muench")
summary(muench2$info)
```

We can plot any model with the `plot()` function.

```{r}
plot(muench2) 
```

### Griffith model

Similarly, we can estimate **Griffith**'s model.

```{r}
gf_model <- polynomial_model(age, pos, neg, type = "Griffith")
plot(gf_model)
```

### Grenfell and Anderson model

And Grenfell and Anderson's model.

```{r}
grf_model <- polynomial_model(age,pos,neg,type = "Grenfell")
plot(grf_model)
```

## Nonlinear models

Refer to `Chapter 6.1.2`

### Farrington model

Use `farrington_model()` to fit a **Farrington**'s model.

```{r, warning=FALSE}
rubella <- rubella_uk_1986_1987
rubella$neg <- rubella$tot - rubella$pos

farrington_md <- farrington_model(
   rubella$age, rubella$pos, rubella$tot,
   start=list(alpha=0.07,beta=0.1,gamma=0.03)
   )
plot(farrington_md)
```

### Weibull model

Use `weibull_model()` to fit a Weibull model.

```{r}
hcv <- hcv_be_2006[order(hcv_be_2006$dur), ]
dur <- hcv$dur
infected <- hcv$seropositive

wb_md <- weibull_model(
   t = dur,
   spos = infected
   )
plot(wb_md) 
```

## Fractional polynomial model

Refer to `Chapter 6.2`

Use `find_best_fp_powers()` to find the powers which gives the lowest deviance score

```{r, warning=FALSE}
hav <- hav_be_1993_1994
best_p <- find_best_fp_powers(
  hav$age, hav$pos, hav$tot,
  p=seq(-2,3,0.1), mc=FALSE, degree=2, link="cloglog"
)
best_p
```

Use `fp_model()` to fit a fractional polynomial model

```{r}
model <- fp_model(
  hav$age, hav$pos, hav$tot,
  p=c(1.5, 1.6), link="cloglog")
plot(model)
```
